Stability of Ensembles of Biological Networks

### 2014-09-12 journal.tex
improve section on stability criteria

\section{Stability analysis of arbitrary systems}

@@

Another form of stability, referred to as structural stability \cite{Smale1967}, requires the determination of whether or not a given dynamical system that is determined to be stable remains stable under a perturbation to one of its defining parameters. This can be formalized as the conditional probability distribution
$$
P(A' \, \textrm{stable} | A \, \textrm{stable})
$$
where $A'$ represents a matrix derived from some perturbation applied to the matrix $A$ given above. In the simple case where $A \in \mathbb{R}^{n \times n}$ the perturbation which involves sampling uniformly over the $a_{ij}$ defining $A$ the value of a particular $a_{ij}$ from a given probability distribution.

\section{Decomposition of stability computations into strongly connected components}
$${\chi(M)'(z) \over \chi(M)(z)} = \tr (zI - M)^{-1}$$
where, $\chi(M)(z) = \det (zI - M)$ is the chararacteristic polynomial of
the matrix $M$ with $z$ a complex variable.  This formula is readily
verified when $M$ is diagonal and has the nice property of being
additive over decompositions of the matrix. Making the change of variables $x=\frac{1}{z}$, we have
$$\tr (I - xM)^n = \sum_{n=0}^\infty x^n \sum_{a \in \Vertex(G)} \sum_{p
\in \Path_n (G)(a,a)} F(p)$$
where $G$ is the graph of the matrix and $\Path_n (G;a,b)$ is the set of
paths of length $n$ from $a$ to $b$.  $(\Vertex(G), \Path(G), \cdot)$ is a category; $F$ is the functor from this category to the multiplicative semigroup of the real numbers which corresponds to the matrix $M$.

@@

\section{Arbitrary system size}
We call a matrix $M$ stable if all its eigenvalues have negative real
part.  Define the polynomial $p(x) = \det(xI + M)$.  Then, since the
roots of $p$ are minus the eigenvalues of $M$, asking that $M$ be
stable is equivalent to asking that the roots of $p$ all have real
part.  Using some good old fashioned invariant theory, we will
formulate conditions for this to happen in terms of the coefficients of
$p$.

@@

Introduce $V$ as the parameter space of all polynomials of degree $n$
with real coefficients.  In other words, $V$ is a real vector space of
dimension $n$ whose coordinates we will identify with the coefficients
$s_k$ of our polynomial.  We now partition this space into regions
corresponding to specific types of polynomials.  Let $R_{abcd}$ be the
subset of $V$ such that $x \in R_{abcd}$ iff the polynomial
corresponding to $x$ has $a$ positive real roots, $b$ negative real
roots, $c$ complex roots with positive real part and $d$ complex roots
with negative real part.  Once we characterize these regions using
invariants, we will have solved the original problem because a matrix
is stable if the point of $V$ corresponding to it lies in a region of
the form $R_{a0c0}$.

@@

Since these quantities are invariant under permutation of the roots
$r_j$, they may be expressed as functions of the coefficients $s_j$.
The explicit expressions may be obtained by such techniques as long
division and determinants.

These quantities are of interest because they describe the boundaries
between the regions introduced above.  Suppose that we have a family
of polynomials whose roots are of one type but turn into another type.
There are three ways this could happen:
\begin{itemize}
\item{1.} A real root could go from being positive to being negative
  or vice versa.  For this to happen, we must pass through an
  intermediate value with a zero root.  For that intermediate value,
  we have $D = 0$.
\item{2.} A pair of real roots could turn into a pair of complex roots
  or vice versa.  For this to happen, we must pass through an
  intermediate value with a repeated root.  For that intermediate
  value, we have $\delta = 0$.
\item{3.} A conjugate pair of complex roots could go from having
  positive real part to having negative real part or vice versa.  For
  this to happen, we must pass through an intermediate value in which
  the real part is zero.  For that intermediate value, the two roots
  are imaginary so, bing conjugates, they sum to zero, hence we have
  $\Delta = 0$.
\end{itemize}
\noindent  Thus, we conclude that the boundaries of our regions
$R_{abcd}$ consist of portions of the surfaces $D=0, \delta=0,
\Delta=0$ and we will characterize them by studying these surfaces.

The simplest case is $n=2$.  There we have the regions $R_{2000},
R_{1100}, R_{0200}, R_{0020}, R_{0002}$ and our invariants are $D =
s_2, \delta = s_1^2 - 4s_2, \Delta = s_1$.  The relation between these
is summarized in the following table:
\begin{align}
R_{2000} \qquad &D > 0, \delta > 0, \Delta > 0 \cr
R_{1100} \qquad &D < 0, \delta > 0 \cr
R_{0200} \qquad &D > 0, \delta > 0, \Delta < 0 \cr
R_{0020} \qquad &\delta < 0, \Delta > 0 \cr
R_{0002} \qquad &\delta < 0, \Delta < 0 \cr
\end{align}

Moving to $n=3$, our invariants now look as follows: $D = s_3, \delta
= 18 s_1 s_2 s_2 + s_1^2 s_2^2 - 4 s_2^3 - 4 s_1^3 s_3 - 27 s_3,
\Delta = s_1 s_2 - s_3$.  However, now we can no longer simply use their
signs to distinguish regions.  For instance, consider the regions
$R_{3000}$ and $R_{1200}$.  These both have $D > 0$ and $\delta > 0$
but to tell them apart, we need to note that the region $\delta > 0$
has two connected components and use some auxiliary equalities to
specify which of the components contains our region.  (Looking at the
sign of $\Delta$ will not help here because, while $\Delta > 0$ in the
region $R_{3000}$, it is also positive for some elements of $R_{1200}$
as well.)

\section{Two component systems}
We call a matrix stable when the real part of all its eigenvalues is negative.  For a $2 \times 2$ matrix, this amounts to the conditions $T < 0$ and $D >
0$ where $T$ and $D$ denote the trace and the determinant.

@@

\begin{figure}[!ht]
\centering
\noindent\includegraphics[width=0.5\columnwidth]{fig/region2x2.pdf}
\caption{{\bf Stability conditions on coefficients of the characteristic polynomial for two component systems.} }
\label{fig:region2x2}
\end{figure}

@@

\begin{figure}[!ht]
\centering
\noindent\includegraphics[width=0.5\columnwidth]{fig/region3x3.pdf}
\caption{{\bf Stability conditions on coefficients of the characteristic polynomial for three component systems.} }
\label{fig:region3x3}
\end{figure}

\bibliographystyle{amsplain}
% \begin{thebibliography}{10}
\bibliography{bib/books,bib/papers}

% \bibitem {A} T. Aoki, \textit{Calcul exponentiel des op\'erateurs
% microdifferentiels d'ordre infini.} I, Ann. Inst. Fourier (Grenoble)
% \textbf{33} (1983), 227--250.

% \bibitem {B} R. Brown, \textit{On a conjecture of Dirichlet},
% Amer. Math. Soc., Providence, RI, 1993.

% \bibitem {D} R. A. DeVore, \textit{Approximation of functions},
% Proc. Sympos. Appl. Math., vol. 36,
% Amer. Math. Soc., Providence, RI, 1986, pp. 34--56.

% \end{thebibliography}

>>>>

\section{Stability analysis of biological networks}

@@

Another form of stability, referred to as structural stability \cite{Smale1967}, requires the determination of whether or not a given dynamical system that is determined to be stable remains stable under a perturbation to one of its defining parameters. This can be formalized as the conditional probability distribution
$$
P(A' \, \textrm{stable}\, \big| \, A \, \textrm{stable})
$$
where $A'$ represents a matrix derived from some perturbation applied to the matrix $A$ given above. In the simple case where $A \in \mathbb{R}^{n \times n}$ the perturbation which involves sampling uniformly over the $a_{ij}$ defining $A$ the value of a particular $a_{ij}$ from a given probability distribution.

\section{System graph and decomposition into strongly connected components}
$${\chi(M)'(z) \over \chi(M)(z)} = \tr (zI - M)^{-1}$$
where, $\chi(M)(z) = \det (zI - M)$ is the chararacteristic polynomial of
the matrix $M$ with $z$ a complex variable.  This formula is readily
verified when $M$ is diagonal and has the nice property of being
additive over decompositions of the matrix. Making the change of variables $x=\frac{1}{z}$, we have
$$\tr (I - xM)^n = \sum_{n=0}^\infty x^n \sum_{a \in \Vertex(G)} \sum_{p
\in \Path_n (G)(a,a)} F(p)$$
where $G$ is the graph of the matrix and $\Path_n (G;a,b)$ is the set of
paths of length $n$ from $a$ to $b$.  $(\Vertex(G), \Path(G), \cdot)$ is a category; $F$ is the functor from this category to the multiplicative semigroup of the real numbers which corresponds to the matrix $M$.

@@

\section{Stability criteria}
We call a matrix $M$ stable if all its eigenvalues have negative real
part.  Define the polynomial $p(x) = \det(xI + M)$.  Then, since the
roots of $p$ are minus the eigenvalues of $M$, asking that $M$ be
stable is equivalent to asking that the roots of $p$ all have real
part.  We will formulate conditions for this to happen in terms of the coefficients of $p$.

@@

Introduce $V$ as the parameter space of all polynomials of degree $n$
with real coefficients.  In other words, $V$ is a real vector space of
dimension $n$ whose coordinates we will identify with the coefficients
$s_k$ of our polynomial.  We now partition this space into regions
corresponding to specific types of polynomials.  Let $R_{abcd}$ be the
subset of $V$ such that $x \in R_{abcd}$ if and only if the polynomial
corresponding to $x$ has $a$ positive real roots, $b$ negative real
roots, $c$ complex roots with positive real part and $d$ complex roots
with negative real part.  Once we characterize these regions using
invariants, we will have solved the original problem because a matrix
is stable if the point of $V$ corresponding to it lies in a region of
the form $R_{a0c0}$.

@@

Since these quantities are invariant under permutation of the roots
$r_j$, they may be expressed as functions of the coefficients $s_j$.
The explicit expressions may be obtained by such techniques as long
division and determinants. These quantities are of interest because they describe the boundaries between the regions introduced above.

The simplest case is $n=2$.  There we have the regions $R_{2000},
R_{1100}, R_{0200}, R_{0020}, R_{0002}$ and our invariants are $D =
s_2, \delta = s_1^2 - 4s_2, \Delta = s_1$.  The relation between these
is summarized by the conditions
\begin{align}
R_{2000} \qquad &D > 0, \delta > 0, \Delta > 0, \cr
R_{1100} \qquad &D < 0, \delta > 0, \cr
R_{0200} \qquad &D > 0, \delta > 0, \Delta < 0, \cr
R_{0020} \qquad &\delta < 0, \Delta > 0, \cr
R_{0002} \qquad &\delta < 0, \Delta < 0. \cr
\end{align}
These regions used for the classification of stability of two-component systems are show in Fig. \ref{fig:region2x2}. The stable regions are given by $R_{2000}$ and $R_{0020}$. All of the other regions represent unstable matrices.

\begin{figure}[!ht]
\centering
\noindent\includegraphics[width=0.5\columnwidth]{fig/region2x2.pdf}
\caption{{\bf Stability conditions on coefficients of the characteristic polynomial for two component systems.} The regions correspond to all possible relationships between the invariants determined by the characteristic polynomial. Colors correspond to the regions given in the main text Green: $R_{2000}$, Red: $R_{1100}$, Yellow: $R_{0200}$, Magenta: $R_{0002}$, Blue: $R_{0020}$.}
\label{fig:region2x2}
\end{figure}

Moving to $n=3$, our invariants now look as follows: $D = s_3, \delta
= 18 s_1 s_2 s_2 + s_1^2 s_2^2 - 4 s_2^3 - 4 s_1^3 s_3 - 27 s_3,
\Delta = s_1 s_2 - s_3$.  However, now we can no longer simply use their
signs to distinguish regions.  For instance, consider the regions
$R_{3000}$ and $R_{1200}$.  These both have $D > 0$ and $\delta > 0$
but to tell them apart, we need to note that the region $\delta > 0$
has two connected components and use the auxiliary condition that $s_1 > 0$ to
specify which of the components contains our region. The regions analogous to Fig. \ref{fig:region2x2} for three component systems where $D=1$ is given in Fig. \ref{fig:region3x3}.
%(Looking at the sign of $\Delta$ will not help here because, while $\Delta > 0$ in the region $R_{3000}$, it is also positive for some elements of $R_{1200}$ as well.)

\begin{figure}[!ht]
\centering
\noindent\includegraphics[width=0.5\columnwidth]{fig/region3x3.pdf}
\caption{{\bf Stability conditions on coefficients of the characteristic polynomial for three component systems.} Colors correspond to the following regions Green: $R_{3000}$, Red: $R_{1200}$, Yellow: $R_{1002}$, Blue: $R_{1020}$. This is the plane $s_3=1$.}
\label{fig:region3x3}
\end{figure}

\section{Two component systems}
For two-component systems having $2 \times 2$ Jacobian matrices, the aforementioned stability criteria result in the conditions $T < 0$ and $D >
0$ where $T$ and $D$ denote the trace and the determinant.

@@

@@

\bibliographystyle{amsplain}
\bibliography{bib/books,bib/papers}

\section{Appendix A}\label{sec:appA}
Suppose that we have a family
of polynomials whose roots are of one type but turn into another type.
There are three ways this could happen:
\begin{itemize}
\item{1.} A real root could go from being positive to being negative
  or vice versa.  For this to happen, we must pass through an
  intermediate value with a zero root.  For that intermediate value,
  we have $D = 0$.
\item{2.} A pair of real roots could turn into a pair of complex roots
  or vice versa.  For this to happen, we must pass through an
  intermediate value with a repeated root.  For that intermediate
  value, we have $\delta = 0$.
\item{3.} A conjugate pair of complex roots could go from having
  positive real part to having negative real part or vice versa.  For
  this to happen, we must pass through an intermediate value in which
  the real part is zero.  For that intermediate value, the two roots
  are imaginary so, bing conjugates, they sum to zero, hence we have
  $\Delta = 0$.
\end{itemize}
\noindent  Thus, we conclude that the boundaries of our regions
$R_{abcd}$ consist of portions of the surfaces $D=0, \delta=0,
\Delta=0$ and we will characterize them by studying these surfaces.
